---
title: "Data 602 - Assignment Four"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---


Ensure you justify all computation and data visualizations with
accompanying code. </br> </br>

**1.** A pharmaceutical company conducted an experiment to compare the
mean amount of time (in days) necessary for a patient to recover from
the effects and complications that follow the onset of the common cold.
In this experiment, thirty-one healthy females between the ages of 25
and 34 years of age were randomly divided into two groups.

In the first group, $n_{VitC} = 15$ females received a daily dose of 500
milligrams of Vitamin C. The second group consisted of
$n_{Placebo} = 16$ females, each of which was given a *placebo*, or
rather a "fake" Vitamin C tablet. The recovery time - the time until the
common cold symptoms had disappeared - for each of the thirty-one
females was observed (in days). The data are provided below.

**Received 500 mgs of Vitamin C:** 6, 7, 7, 7, 8, 7, 7, 8, 7, 8, 10, 6,
8, 5, 6

**Received a placebo:** 10, 12, 8, 6, 9, 8, 11, 9, 11, 8, 12, 11, 9, 8,
10, 9

(a) Do these data indicate that the recovery time is quicker with
    Vitamin C than without? Carry out the appropriate statistical
    investigation with a permutation test. Ensure you show "where" your
    permutation test statistic lies on the distribution. Report your
    (empirical) $P$-value and your statistical inference. For the number
    of permutation tests, use 2999.

    ```{r}
    # Received 500 mgs of Vitamin C
    vit_c <- c(6, 7, 7, 7, 8, 7, 7, 8, 7, 8, 10, 6, 8, 5, 6)

    # Received a placebo
    placebo <- c(10, 12, 8, 6, 9, 8, 11, 9, 11, 8, 12, 11, 9, 8, 10, 9)

    # Calculate observed difference in mean recovery times
    diff <- mean(vit_c) - mean(placebo)
    print(paste("Observed difference in mean recovery times:", diff))

    # Combined
    both <- c(vit_c, placebo)

    #2999 permutations
    set.seed(123)  
    perm_diffs <- numeric(2999)
    for (i in 1:2999) {
      
      vit_c_perm <- sample(both, 15)
      placebo_perm <- setdiff(both, vit_c_perm)
      
      # Calculate difference in mean recovery times for this permutation
      perm_diff <- mean(vit_c_perm) - mean(placebo_perm)
      perm_diffs[i] <- perm_diff
    }

    # Calculate empirical P-value
    p_value <- mean(perm_diffs <= diff)
    print(paste("P-value:", p_value))
    ```

(b) Re-test your statistical hypothesis in part (a) using the $t$-test.
    In doing so, state any assumptions about these data or conditions
    you are imposing on these data and conduct the necessary diagnostics
    to either confirm or refute such assumptions. Ensure you provide
    both the $P$-value and its interpretation related to these data.

    ```{r}
    # Perform t-test
    t_stat <- t.test(vit_c, placebo)$statistic
    t_p <- t.test(vit_c, placebo)$p.value
    print(paste("t-test statistic:", t_stat))
    print(paste("t-test P-value:", t_p))

    ```

</br>

**2.** A group of patients with a binge-eating disorder were randomly
assigned to take either the experimental drug fluvoxamine or the placebo
in a nine-week-long, double-blinded clinical trial. At the end of the
trial the condition of each patient was classified into one of four
categories: no response, moderate response, marked response, or
remission. The table below shows a cross-classification, or contingency
table, of these data. $$
\begin{array}{lcccc}
             & \text{No Response}  & \text{Moderate Response}  & \text{Marked Response}  & \text{Remission}  \\
             \hline
\text{Fluvoxamine}  &   15         &     7                     &    3                    &   15              \\
\text{Placebo}      &   22         &     7                     &    3                    &   11              \\
\end{array}
$$

Do these data provide statistically significant evidence to conclude
that there is an association between the type of treatment received and
a patient's response?

Answer

-   Null hypothesis  no relation between the type of treatment
    received and a patient's response.

-   Alternative hypothesis Relation between the type of treatment
    received and a patient's response.

    ```{r}

    df <- data.frame(treatment = c(rep("Fluvoxamine", 40), rep("Placebo", 43)),
                     response = c(rep("No Response", 15), rep("Moderate Response", 7), rep("Marked Response", 3), rep("Remission", 15),
                                  rep("No Response", 22), rep("Moderate Response", 7), rep("Marked Response", 3), rep("Remission", 11)))

    table <- table(df$treatment, df$response)
    table
    ```

    ```{r}
    # Perform the chi-squared test
    chisq_test_result <- chisq.test(table)
    # in above line of code i had " Warning in chisq.test(table) :Chi-squared approximation may be incorrect--- so stimulate the p value"
    chisq_test_result

    ```

    ```{r}
    # to have the exact p value fissher is used 
    fisher.test(table)
    ```

</br>

**3.** A study conducted by Youmans and
Jee[1](Fudging%20the%20Numbers:%20Distributing%20Chocolate%20Influences%20Student)
looked at students who took a research methods class in psychology. Two
lecture sections of the same course were taught by the same instructor.
Students in each lecture section were required to register in one of two
Friday afternoon discussion sections, or a tutorial. During the ninth
week of instruction, all students were asked to fill out an informal
midsemester evaluation in their Friday afternoon discussion session.

```         
Evaluations in an Undergraduate Course, *Teaching of Psychology*,
Volume 34, Number 7, pgs.45 - 247, 2007.
```

In half of the Friday afternoon discussion sections, the experimenter
(not the professor/instructor!) passed around a bag containing small
bars of chocolate that he simply had 'left-over' and 'wanted to get rid
of' so that students would not think the chocolate was a gift from their
professor/instructor. The evaluation was a survey that contained nine
questions. For each question the student provided a rating from 1 (very
poor) to 5 (excellent).

Question 9 posed the following:

</br>

<center>**Compared to other instructors you have had at this university,
this instrutor is:'' 1 (very poor) to 5 (excellent)**</center>

</br>

Compliments to the authors, they have provided me with the data in this
particular study. In the
[file](https://github.com/Statman44/Data602/blob/main/chocnochocratings.csv)
you will find the raw data providing the student responses for Question
9 (**Q9**) and the overall average response for Questions 1 through 9
(**Overall**) for each student. *Note: To easily read this into your R
session, select the "Raw" button prior, then use the resulting URL.*

(a) Pertaining to Question 9: Do these data suggest there is a treatment
    effect? Test the existence of a treatment effect using a permutation
    test. If a treatment effect is discovered, explain its meaning in
    the context of these data. Carry out a permutation test, using 1999
    iterations.\
    </br> (You will have to carefully consider "what a treatment effect"
    is in this scenario.)

    ```{r}
    # carring out teh permuation test using 1999 iterations.
    ## Load dataset
    data<- read.csv(file.choose(),header=T, na.strings = c("","NA"))
    head(data)
    ```

    ```{r}
    names(data)
    ```

    ```{r}
    chocolate <- data$Q9[data$GroupName == "Chocolate"]
    No_chocolate <- data$Q9[data$GroupName == "NOChoc"]

    #difference in means
    diff <- mean(chocolate, na.rm= TRUE) - mean(No_chocolate, na.rm= TRUE)
    diff

    set.seed(123) 
    permutation_diff <- replicate(1999, diff(sample(data$Q9)), simplify = FALSE)


    p_value <- sum(abs(unlist(permutation_diff)) >= abs(diff)) / 1999

    # Compare the p-value
    if (p_value < 0.05) {
      cat("Reject the null hypothesis.\n")
    } else {
      cat("Fail to reject the null hypothesis.\n")
    }

    # Print the p-value
    cat("p-value:", p_value, "\n")
    ```

(b) Consider the variable **Overall**. Is there a treatment effect with
    respect to the professor's overall rating as a teacher? Apply the
    $t$-test to these data. Interpret the meaning of the $P$-value.

    ```{r}

    Overall_Chocolate <- data$Overall[data$GroupName == "Chocolate"]
    Overall_NOChoc <- data$Overall[data$GroupName == "NOChoc"]

    # Perform the two-sample t-test
    t_test_result <- t.test(Overall_Chocolate, Overall_NOChoc)

    # Print the t-test result
    print(t_test_result)
    ```

(c) Consider the test suggested in part (a). Why would a $t$-test not be
    a recommended statistical method to carry out the test in part (a)?
    Explain your answer in a few sentences

    **Answer:**

    **The permutation test is specifically designed to test for
    differences in means between two groups, making it a more suitable
    choice for this particular problem.**

</br> </br>

**4.** The Capital Asset Price Model (CAPM) is a financial model that
attempts to predict the rate of return on a financial instrument, such
as a common stock, in such a way that it is linearly related to the rate
of return on the overal market. Specifically

$$
R_{StockA,i} = \beta_{0} + \beta_{1}R_{Market, i} + e_{i}   
$$

(Note: To align the notation with a stock's \`\`beta'', $\beta_{0} = A$
and $\beta_{1} = B$)

You are to study the relationship between the two variables and estimate
the above model:

$R_{SUNCOR, i}$ - rate of return on Stock A for month $i$,
$i = 1, 2, \cdots, 59$.

$R_{Market, i}$ - market rate of return for month $i$,
$i = 1, 2, \cdots, 59$.

$\beta_{1}$ represent's the stocks \`beta' value, or its *systematic
risk*. It measure's the stocks volatility related to the market
volatility. $\beta_{0}$ represents the risk-free interest rate.

The data appeaing in the
[file](https://github.com/Statman44/Data602/blob/main/bondsdata.csv)
contains the data on Suncor's rate of return and the Toronto Composite
Index rate of return for 59 randomly selected months.

Therefore $R_{SUNCOR, i}$ represents the monthly rate of return for a
common share of Suncor stock; $R_{TSE, i}$ represents the monthly rate
of return (increase or decrease) of the TSE Index for the same month,
month $i$. The first column in this data file contains the monthly rate
of return on Suncor stock; the second column contains the monthly rate
of return on the TSE index for the *same month*.

Read this data into R Studio and answer the questions posed below.

```{r}
capmdata.df = read.csv("https://raw.githubusercontent.com/Statman44/Data602/main/capm.csv")
head(campdata.df, 3) 
```

(a) Appropriately visualize these data. What can you infer from this
    visualization? Provide a brief commentary.

(b) Estimate the model above.

(c) In the context of these data, interpret the meaning of your
    estimates of the estimates $\widehat{\beta}_{0}$ and
    $\widehat{\beta}_{1}$, in the context of these data.

(d) Refer to your answer in (b) In a certain month, the rate of return
    on the TSE Index was 4%. Predict the rate of return on Suncor stock
    for the same month.

(e) Think about the conditions of this model *in the context* of these
    data. Create the visualizations that inspect each of the two
    conditions and provide commentary that addresses the validity (or
    invalidity) of each.

(f) From these data, can you infer that the monthly rate of return of
    Suncor stock can be expressed as a positive linear function of the
    monthly rate of return of the TSE Index? State your statistical
    hypotheses, compute (and report) both the test statistic and the
    $P$-value and provide your decision.

(g) Compute a 95% confidence interval for $\beta_{1}$, then interpret
    its meaning in the context of these data.

(h) Compute a 95% confidence interval for the mean monthly rate of
    return of Suncor stock when the TSE has a monthly rate of return of
    3%.

(i) In a month of September, the TSE Index had a rate of return of
    1.16%. With 95% confidence, compute the September rate of return for
    Suncor stock.

(j) Recall the Bootstrap Method. From these data, use the bootstrap
    method to create a 95% confidence interval for mean monthly rate of
    return of Suncor stock when the TSE has a monthly rate of return of
    3%. Compare your result to your result in part (h). Use 1000
    iterations for your bootstrap. *Carefully* consider how you would
    resample bivariate data points $(x_{TSE, i}, y_{Suncor, i})$.

</br> </br>

```{r}
library(ggplot2)
library(dplyr)
library(boot)

capmdata.df <- read.csv("https://raw.githubusercontent.com/Statman44/Data602/main/capm.csv")
head(capmdata.df, 3)




```

```{r}
colnames(capmdata.df) <- c("Suncor_Return", "TSE_Return")
ggplot(capmdata.df, aes(x = TSE_Return, y = Suncor_Return)) +
  geom_point(color = 'blue') +
  geom_smooth(method = "lm", col = "red", se = FALSE) +
  labs(title = "Suncor Stock Return vs TSE Index Return",
       x = "TSE Index Return (%)",
       y = "Suncor Stock Return (%)")


### 2. Estimate the Model
capm_model <- lm(Suncor_Return ~ TSE_Return, data = capmdata.df)
summary(capm_model)

predict_return <- predict(capm_model, newdata = data.frame(TSE_Return = 0.04))
cat("Predicted Suncor stock return for a TSE return of 4%:", predict_return, "\n")

ggplot(capm_model, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, col = "red") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals")

# Normal Q-Q Plot (for checking normality of residuals)
ggplot(capm_model, aes(sample = .stdresid)) +
  stat_qq() +
  stat_qq_line(col = "red") +
  labs(title = "Normal Q-Q Plot of Residuals")



# Test statistic and p-value are available from the model summary
cat("Test statistic (t-value for beta_1):", summary(capm_model)$coefficients[2, "t value"], "\n")
cat("P-value:", summary(capm_model)$coefficients[2, "Pr(>|t|)"], "\n")



### 6. 95% Confidence Interval for beta_1 ###
confint(capm_model, "TSE_Return", level = 0.95)



### 7. 95% CI for Mean Suncor Return when TSE Return = 3% ###
newdata <- data.frame(TSE_Return = 0.03)
predict(capm_model, newdata, interval = "confidence", level = 0.95)


### 8. 95% CI for September (TSE Return = 1.16%) ###
newdata_september <- data.frame(TSE_Return = 0.0116)
predict(capm_model, newdata_september, interval = "confidence", level = 0.95)


### 9. Bootstrap for 95% CI when TSE Return = 3% ###
# Bootstrap function for resampling
boot_fn <- function(data, indices) {
  boot_data <- data[indices, ]
  boot_model <- lm(Suncor_Return ~ TSE_Return, data = boot_data)
  predict(boot_model, newdata = data.frame(TSE_Return = 0.03))
}

# Perform bootstrap with 1000 iterations
set.seed(123)
bootstrap_results <- boot(capmdata.df, statistic = boot_fn, R = 1000)

# Calculate 95% confidence interval using bootstrap results
boot.ci(bootstrap_results, type = "perc")

```
