---
title: "Data602_Project"
author: "Venkateshwaran"
date: "2024-10-11"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 3:Analysis of Fire Size Vs Response & Extinguished Time 

## Introduction
Analyze the relationship between the Fire Size against the Response /Extinguished Time

### Response Time
Duration between the detection of a forest fire and the arrival of firefighting resources at the scene.

### Extinguished Time
Duration between the detection of a forest fire the moment when the fire is completely put out and no longer poses a threat.


```{r}
library(mosaic)
library(dplyr)
library(ggplot2)
```

## Data Cleaning and Transformation

### Reading the file and converting the blank into NA

```{r}
data1<-read.csv(
"C:\\Users\\Venkateshwaran\\OneDrive\\Desktop\\Data602_Project\\Revised_Project\\fp-historical-wildfire-data-2006-2023_rawdata.csv",
na.strings=c("","NA"))
```

### Step 1: Identifying the targetted Variable:

Below are the list of variables which we are focusing to continue our statistical analysis for this problem statement

- Fire_Size ( in Hectares)
- fire_start_date 
- discovered date
- ex_fs_date -(Extinguished time)
- assessment_datetime-(Actual Response Starts)


### Step 2: Mapping the NA assessment_datetime,ex_fs_date

Identified the blank dates with the next earliest date for fire_start_date with Discovered_date, assessment date with the arrival date in the fire control date,bh_fs_date. Main Objective of this transformation is not to eliminate the data which will be really supporting our regression


```{r}
data2 <- data1 %>% 
   mutate(discovered_date = ifelse(discovered_date == 'NA', reported_date, discovered_date),
          ex_fs_date = ifelse(ex_fs_date == 'NA', bh_fs_date, ex_fs_date),
          fire_start_date = ifelse(fire_start_date == 'NA', discovered_date, fire_start_date),
          assessment_datetime = ifelse(assessment_datetime == 'NA', ia_arrival_at_fire_date, assessment_datetime),
          discovered_date = ifelse(is.na(discovered_date), reported_date, discovered_date),
          ex_fs_date = ifelse(is.na(ex_fs_date), bh_fs_date, ex_fs_date),
          fire_start_date = ifelse(is.na(fire_start_date), discovered_date, fire_start_date),
          assessment_datetime = ifelse(is.na(assessment_datetime), ia_arrival_at_fire_date, assessment_datetime))
```

### Step 3: Converting the date variables into date format

```{r}
data2$ex_fs_date <- as.POSIXct(data2$ex_fs_date, format = "%Y-%m-%d %H:%M:%S", tz = "America/Edmonton")
data2$discovered_date <- as.POSIXct(data2$discovered_date, format = "%Y-%m-%d %H:%M:%S", tz = "America/Edmonton")
data2$fire_start_date <- as.POSIXct(data2$fire_start_date, format = "%Y-%m-%d %H:%M:%S", tz = "America/Edmonton")
data2$assessment_datetime <- as.POSIXct(data2$assessment_datetime, format = "%Y-%m-%d %H:%M:%S", tz = "America/Edmonton")
# Check the classes after conversion
class(data2$ex_fs_date)
class(data2$fire_start_date)
```

### Step 3: Calculating the response and extinguished time in minutes

- Responsetime = assessment_datetime-fire_start_date
- Exttime= ex_fs_date-fire_start_date

```{r}
# Calculate response time and ext time in minutes
data2$Responsetime <- as.numeric(difftime(data2$assessment_datetime, data2$fire_start_date, units = "mins"))
data2$Exttime <- as.numeric(difftime(data2$ex_fs_date, data2$fire_start_date, units = "mins"))
```

### Step 4: Logarthimic Transformation

- Performed a log linear transformation on the dependent variables(current_size,Responsetime,Exttime)

```{r}
data2$current_size=log10(data2$current_size)
data2$Responsetime=log10(data2$Responsetime)
data2$Exttime=log10(data2$Exttime)
```

### Step 5: Filtering Variables

-Filtering Out the required variables to continue with the further regression analysis of the size and the times which we are targetting to produce the correlation.`

```{r}
data_filtered<- data2 %>%
  select(fire_number, fire_year, current_size, Responsetime, Exttime)
head(data_filtered)
```

### Step 6: Calulating the mean response time and extinguished time

- To continue analysis of the newly derived variables and conducting analysis, we performed grouping the time with respect to the affected size and storing it into different data frames
- result1 - (Response Time)
- result2 - (Extinguished Time)

Below are the cleaned data set which are going to be in our further regression Analysis


```{r}

result1 <- data2 %>%
  group_by(current_size) %>%
  summarise(
    meanresponsetime = mean(Responsetime),
  )

result1 <- result1 %>%
  filter(!is.na(meanresponsetime), !is.infinite(meanresponsetime),
         !is.na(current_size), !is.infinite(current_size))


result2 <- data2 %>%
  group_by(current_size) %>%
  summarise(
    meanExttime = mean(Exttime),
  )%>%
  arrange(meanExttime)

result2 <- result2 %>%
  filter(!is.na(meanExttime), !is.infinite(meanExttime),
         !is.na(current_size), !is.infinite(current_size))
cat("Data Frame for Response Time")

head(result1)
summary(result1)
cat(".\n")
cat("Data Frame for Extinquished Time")
head(result2)
summary(result2)
```

# Regression Analysis

## Plotted the scattered plots to validate the density of the variables being analyzed

```{r}
ggplot1= (ggplot(result1, aes(x=meanresponsetime, y = current_size))+
  geom_point (color = "red") +
  stat_smooth(method = "lm", formula = y ~ x, geom = "smooth")+
  labs (title = "Scatter plot of Average Response Time and current_size",
        x = "Response Time", y ="Average current_size"))
ggplot1
```

## Inference:
-Positive linear trend. Response time increases, the current size also slightly increases, though the relationship seems weak (the slope of the line is small).

-Data points (in red) are densely scattered, with many points clustering around lower response times and a current size. However, there are a few points that deviate significantly from this dense cluster, especially for higher response times and current sizes.

```{r}
ggplot2 =(ggplot(result2, aes(x=meanExttime, y = current_size))+
  geom_point (color = "red") +
  stat_smooth(method = "lm", formula = y ~ x, geom = "smooth")+
  labs (title = "Scatter plot of Average Extinguished time and Average current_size",
        x = "Extinguished time", y ="Average current_size"))
ggplot2
```
## *Inference:*

-Strong Positive Trend
-Slope of the line is noticeably steeper compared to the response time plot, indicating that changes in extinguished time have a more pronounced effect on the average current size.



# Generating Models and Intreprating the Model Summary with Hypothesis Test

### Mean Response Time

-Null Hypothesis (H0): There is no relationship between the Response time and Average current_size.
-Alternative Hypothesis (Ha): There is a relationship between the Response time and Average current_size.

```{r}
model_clean1 = lm(current_size ~ meanresponsetime, data = result1)
summary(model_clean1)
```

## Interpretation: 
p-value of 0.03362 relationship between meanresponsetime and current_size is statistically significant at the 5% significance level (p < 0.05)

## Model Fit: 
R-squared value of 0.002397 indicates a Low Fit.



### Extinquished Time

-Null Hypothesis (H0): There is no relationship between the Extinguished time and Average current_size.
-Alternative Hypothesis (Ha): There is a relationship between the Extinguished time and Average current_size.

```{r}
model_clean2 = lm(current_size ~ meanExttime, data = result2)
summary(model_clean2)
```


## Interpretation:
p-value of 2.2e-16  is extremely low, indicating strong evidence against the null hypothesis means that Response time has significant effect on the Size of the forest impact

## Model Fit:  
R-squared: 0.5386 of the variance in current_size which indicates a moderate fit



# Analyzing the Normal Distribution - QQ Plot for Response Time

```{r}
qqnorm(model_clean1$residuals)
qqline(model_clean1$residuals, col = "red")
```
## Interpretation: 
- Model appears to follow a normal distribution reasonably well. 
- Some deviations at the extreme ends (in the tails) may indicate slight departures from normality, possibly showing mild outliers or heavier tails than a normal distribution



# Analyzing the Normal Distribution - QQ Plot for Extinquished Time

```{r}
qqnorm(model_clean2$residuals)
qqline(model_clean2$residuals, col = "red")
```

## Interpretation:

- Model is approximately normal
- Model may have heavier tails or more extreme values than would be expected in a normal distribution- When the extinguished time increases there will be huge impact on the affected area


# Analyzing the Normal Distribution - Residual Plot for Response Time

```{r}
plot(model_clean1$fitted.values, model_clean1$residuals)
abline(h = 0, col = "red")
```
## Inference:
-Residuals seem to be randomly scattered around the horizontal line at zero though there are many outliers in the data set



# Analyzing the Normal Distribution - Residual Plot for Extinguished Time

```{r}
plot(model_clean2$fitted.values, model_clean2$residuals)
abline(h = 0, col = "red")
```
## Inference:
-Residuals are centered around zero, which is a good sign for a well-fitted model and  shows the strong relationship between time and size



# Conclusion

-Overall, Models of Response and Extinguished time has a significant relationship between the affected area - size.
-However, the Extinguished Time Model which shows the strong relationship with the relatively low p-value
-Hence improving the Response and extinguished time of a forest fire involving with following strategies ideally impact the size of the forest being affected
-Early detection,Rapid response,Effective firefighting techniques,Training and Coordination & Public awareness 




